Visual narratives of the COVID-19 pandemic

### 2022-10-27 JDSSV_manuscript_submitted.Rnw
copyediting_secondRound

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards\citep{kochWelcomeRevolutionCOVID192021}.

@@

Proportional symbol maps, see \Cref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. 
The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight\citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
In addition, the visual impression of any map based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.

@@

Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as colour). Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.  While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19\citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information. This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes. 
A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.

@@

Throughout the pandemic, both in choropleth maps and time series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

@@

Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}. 
The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely. 
Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.

@@

While traditionally, time series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic. In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity; for example, John Burn-Murdoch's decision to index Financial Times COVID graphs to the days since the $N$th case or death. This decision is a small modification of the basic time series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "... that there's an inevitability about how coronavirus spreads... even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically."\citep{hannenCoronavirusTrajectoryTracker2020a} By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time\citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.

@@

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube\citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}. While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way. A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualisations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.

A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}. The visualisation comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one. The form, animation, and auditory experience are all indicative of an emotional appeal \citep{kostelnick_re-emergence_2016,dignazio_data_2020}; with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.

@@

Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies\citep{weissThese12Graphs2020}. \Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases. Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear; the reader is supposed to conclude that cases and masks are not associated. These charts were so common on social media that media outlets ran stories identifying them as misinformation \citep{reutersstaffFactCheckMask2020}; clearly, comparative charts were effectively deployed to misinform as well as to inform.

@@

\section{To log or not to log}\label{sec:logs}
As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult. When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
% Discussion of how this section's conclusions change with the pandemic

@@

%% -- Optional special unnumbered sections -------------------------------------
\newpage
\section*{Computational Details}
Graphics included in this paper not sourced from media outlets were created using data published by the New York Times at \url{https://github.com/nytimes/covid-19-data/}. Graphics were created using ggplot2\citep{ggplot2} and \proglang{R}~4.1.2. \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

>>>>

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.

@@

Proportional symbol maps, see \Cref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. 
The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight \citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
In addition, the visual impression of any map based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.

@@

Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as colour). Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.  While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19 \citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information. This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes. 
A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.

@@

Throughout the pandemic, both in choropleth maps and time series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed \citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

@@

Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}. 
The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely. 
Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts \citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change \citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.

@@

While traditionally, time series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic. In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity; for example, John Burn-Murdoch's decision to index Financial Times COVID graphs to the days since the $N$th case or death. This decision is a small modification of the basic time series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "... that there's an inevitability about how coronavirus spreads... even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically" \citep{hannenCoronavirusTrajectoryTracker2020a}. By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time \citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.

@@

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube \citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names \citep{barryRemembering1000002020}. While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way. A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualisations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.

A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time \citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}. The visualisation comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one. The form, animation, and auditory experience are all indicative of an emotional appeal \citep{kostelnick_re-emergence_2016,dignazio_data_2020}; with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.

@@

Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies \citep{weissThese12Graphs2020}. \Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases. Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear; the reader is supposed to conclude that cases and masks are not associated. These charts were so common on social media that media outlets ran stories identifying them as misinformation \citep{reutersstaffFactCheckMask2020}; clearly, comparative charts were effectively deployed to misinform as well as to inform.

@@

\section{To log or not to log}\label{sec:logs}
As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly \citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances \citep{hecklerStudentAccuracyReading2013} is difficult. When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
% Discussion of how this section's conclusions change with the pandemic

@@

%% -- Optional special unnumbered sections -------------------------------------
\newpage
\section*{Computational Details}
Graphics included in this paper not sourced from media outlets were created using data published by the New York Times at \url{https://github.com/nytimes/covid-19-data/}. Graphics were created using ggplot2 \citep{ggplot2} and \proglang{R}~4.1.2. \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.
